Type theories as quotient inductive-recursive types

### 2025-02-19 tex/main.tex
Polish 3.4 (currently 15.1 pages)

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
%
In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by the reduction
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
\end{alignat*}
Yet, this naive change breaks the local confluence!
The problem is that there is now additional interaction between types and terms. Observe that the local peak
\[
  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
\]
cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
%
To repair the local confluence, we add a simultaneously defined function $\subTm{\blank}{\blank}$ for term substitution apart from the explicit term substitution $\sub{\blank}{\blank}$ as follows
\begin{alignat*}{3}
  \subTm{\blank}{\blank} & : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma}) \\
\subTm{t}{\idS}          & \reduce t \\
\subTm{t}{\sigma ; \tau} & \reduce \subTm{(\subTm{t}{\tau})}{\sigma} \\
\subTm{t}{\pi_1(\sigma, s)} & \reduce \subTm{t}{\sigma} \\
\subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{(\subTm{t}{\pi_1\tau})}{\sigma} \\
\subTm{t}{\sigma}        & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
\end{alignat*}
which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
The remaining cases reduce to the explicit term substitution $\sub{t}{\sigma}$.
Note that $\subTm{\blank}{\blank}$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
\begin{remark}
  The function $\subTm{\blank}{\blank}$ can be understood as a \emph{smart constructor} for the explicit term substitution $\sub{\blank}{\blank}$, a well-known idiom in functional programming, since the function $\subTm{\blank}{\blank}$ (mostly) builds an explicit substitution but computes certain cases away.
\end{remark}

The equality constructor $[]\El$ is now accordingly replaced by the reduction
\begin{align}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\subTm{u}{\sigma}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution. The proof is similar to the proof of \cref{prop:local-confluence-1}, and can be found in \cref{appendix:proofs}.
%\begin{proposition}[Termination]
%  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
%  \danger
%\end{proposition}
%\begin{proof}
%  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
%\end{proof}
\begin{restatable}{proposition}{propConfluenceTwo}\label{prop:local-confluence-2}
  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
  \danger
\end{restatable}

@@

\begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$.
\end{restatable}

Finally, we have to show that both $\subTy{\blank}{\blank}$ and $\subTm{\blank}{\blank}$ are coherent.
%
\begin{restatable}{proposition}{propCoherence}\label{prop:coherence-2}
  For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, the identities
  \danger
  \[
    \subTy{A}{\sigma} = \subTy{A}{\tau}
    \quad\text{and}\quad
    \subTm{t}{\sigma} = \subTm{t}{\tau}
  \]
  hold for any type $A$ and any term $t$.
\end{restatable}

\begin{remark}
We have seen that a naive extension with a universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this propagation can be achieved by turning a term involving a substitution $\sigma$ into a recursion.
\end{remark}

@@

We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, where the context of its codomain is extended with the domain type.
For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to adding the following constructors for the type formation rule, the term introduction/elimination rules:
\begin{alignat*}{3}
  \Pi     &: \implicit{\Gamma}           \; && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
  \lambda &: \implicit{\Gamma, A, B}     \; && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
  \mathsf{app} &: \implicit{\Gamma, A, B}\; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
the computation rules
\begin{alignat*}{5}
  \Pi\beta           & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t) && =^{\Tm\;\Gamma\;A} && t \\
  \Pi\eta            & : \implicit{\Gamma, A, B, t}\; && t                          && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
\end{alignat*}
and the structural rules for substitution
\begin{alignat}{5}
  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && \subTy{(\Pi\;A\;B)}{\sigma} && =^{\Ty\;\Gamma} && \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) \label{eq:def-type-subst-8} \\
  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && \subTm{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
\end{alignat}
where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_T}{\color{black}\pi_2\idS}})$.

For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are strict.
we do not need any transport since $[;]_T$ and $[]\Pi$ are strict.
However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $\subTy{\blank}{\blank}$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
For example, $\subTy{(\Pi\;A\;B)}{\idS}$ reduces to $\Pi\;A\;(\subTy{B}{\idS \uparrow A})$ where $\idS \uparrow A$ is propositionally but not strictly equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow \subTy{(\Pi\;A\;B)}{\idS} \reduce \Pi\;A\;(\subTy{B}{\idS \uparrow A})$ cannot be joined.

To retain the local confluence, we turn $\sigma \uparrow A$ into a simultaneously defined function:
\begin{alignat}{3}
  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, \subTy{A}{\sigma})\, (\Delta, A) \label{eq:type-of-lifting} \\
\idS                \uparrow A & \reduce \idS \nonumber \\
(\sigma ; \tau)     \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \nonumber \\
\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \nonumber \\
\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow \sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A)\nonumber \\
\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwise} \nonumber
\end{alignat}
where $\sigma^+ \defeq (\pi_1\idS ; \sigma), \pi_2\idS$.
Then, we add another clause for type substitution and an equality constructor for term substitution without transport:
\begin{alignat}{5}
                     & \subTy{(\Pi\,A\,B)}{\sigma} && \reduce \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) \label{eq:def-type-subst-9} \\
  \mathsf{[]\lambda} :{} & \sub{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
\end{alignat}

@@

As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ but the same argument still applies:
\begin{proposition} \label{prop:correctness-3}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
\end{proposition}

Finally, we show that these simultaneously defined functions are coherent.
\begin{restatable}{proposition}{propCoherenceUparrow} %[Coherence]
    For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, we have
    \danger
    \[
      \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, \subTy{A}{\gamma})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A
      \qquad\text{and}\qquad
      \subTm{t}{\sigma} =^{\lam{\gamma}\Tm\;\Gamma\;(\subTy{A}{\gamma})}_{\alert{\cc}} \subTm{t}{\tau},
    \]
    for any type $A$ and any term $t$.
\end{restatable}

@@

We also need to prove the coherence for type substitution, i.e., that $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
\[
  \subTy{(\Pi\;A\;B)}{\sigma} = \Pi\;(\subTy{A}{\sigma})\;(\subTy{B}{\sigma\uparrow A}) \qquad
  \subTy{(\Pi\;A\;B)}{\tau} = \Pi\;(\subTy{A}{\tau})\;(\subTy{B}{\tau\uparrow A})
\]
but the induction hypothesis does not give us $\subTy{B}{\sigma\uparrow A} = \subTy{B}{\tau\uparrow A}$.
Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
Then, for each constructor $\cc : \sigma = \tau$, we show the identity
$\subTy{A}{\sigma \mathop{\upuparrows} \Xi} = \subTy{A}{\tau \mathop{\upuparrows} \Xi}$
for arbitrary $\Xi$, subsuming the special case $\subTy{A}{\sigma} = \subTy{A}{\tau}$ when $\Xi$ is empty.

The type of telescopes over a `base' context $\Gamma$ $\Tel : \Ctx \to \Set$ are defined inductive-recursively simultaneously with a concatenation $\blank \dplus \blank$ as follows.

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \Tel               & :                   && \Ctx \to \Set                                     \\
  \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
  \Gamma & \dplus \emptytel && \reduce \Gamma \\
  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
Equipped with telescopes, we can now generalise single lifting with type \eqref{eq:type-of-lifting} to multiple lifting with type
\[
\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
\to \Sub\;(\Gamma \dplus (\subTel{\Xi}{\sigma}))\;(\Delta \dplus \Xi)
\]
which requires us to also simultaneously define telescope substitution $\subTel{\blank}{\blank} :  \Tel\;\Delta \to \Sub\;\Gamma\;\Delta \to \Tel\;\Gamma$.

@@

\begin{minipage}{.5\textwidth}
\begin{alignat*}{3}
  & \subTel{\Xi}{\idS}       \Xi && \reduce \Xi \\
  & \subTel{\Xi}{\sigma ; \tau} && \reduce \subTel{(\subTel{\Xi}{\tau})}{\sigma}
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & \idS          && \upuparrows \Xi && \reduce \idS \\
  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
\end{alignat*}
\end{minipage}\vskip\belowdisplayskip\noindent
%whose confluence property is easy to see.

>>>>

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
%
In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by the reduction
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
\end{alignat*}
Yet, this naive change breaks local confluence!
The problem is that there is now additional interaction between types and terms.
% Observe that
The local peak
\(
  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
\)
cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
%
To repair local confluence, we add a simultaneously defined term substitution function $\subTm{\blank}{\blank} : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma})$, different from the constructor $\sub{\blank}{\blank}$, with the following reductions:
\begin{alignat*}{3}
%  \subTm{\blank}{\blank} & : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma}) \\
\subTm{t}{\idS}          & \reduce t \\
\subTm{t}{\sigma ; \tau} & \reduce \subTm{(\subTm{t}{\tau})}{\sigma} \\
\subTm{t}{\pi_1(\sigma, s)} & \reduce \subTm{t}{\sigma} \\
\subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{(\subTm{t}{\pi_1\tau})}{\sigma} \\
\subTm{t}{\sigma}        & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
\end{alignat*}
This reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from types to terms.
The remaining ``default'' cases reduce to the constructor $\sub{t}{\sigma}$.
Note that $\subTm{\blank}{\blank}$ can be defined with the standard first-match pattern matching semantics, since there are no overlapping patterns.
\begin{remark}
  The function $\subTm{\blank}{\blank}$ can be understood as a \emph{smart constructor} for the explicit term substitution $\sub{\blank}{\blank}$, a well-known idiom in functional programming: the function $\subTm{\blank}{\blank}$ computes certain cases away, but mostly reduces to the non-smart constructor $\sub{\blank}{\blank}$.
\end{remark}

The equality constructor $[]\El$ is now accordingly replaced by the reduction
\begin{align}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\subTm{u}{\sigma}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution. The proof is similar to the proof of \cref{prop:local-confluence-1}, and can be found in \cref{appendix:proofs}.
%\begin{proposition}[Termination]
%  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
%  \danger
%\end{proposition}
%\begin{proof}
%  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
%\end{proof}
\begin{restatable}{proposition}{propConfluenceTwo}\label{prop:local-confluence-2}
  \danger
  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
\end{restatable}

@@

\begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
\end{restatable}

Finally, we have to show that both $\subTy{\blank}{\blank}$ and $\subTm{\blank}{\blank}$ are coherent.
%
\begin{restatable}{proposition}{propCoherence}\label{prop:coherence-2}
  For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, the identities
  \danger
    $\subTy{A}{\sigma} = \subTy{A}{\tau}$
    and
    $\subTm{t}{\sigma} = \subTm{t}{\tau}$
  hold for any type $A$ and any term $t$.
\end{restatable}

\begin{remark}
We have seen that a naive addition of a universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with existing clauses.
In general, every term that occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this propagation can be achieved by turning a constructor involving a substitution into a recursive function.
\end{remark}

@@

We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types. % --- the context of the codomain is extended with the domain type.
For a QIIT definition~\cite{Altenkirch2016a}, this amounts to adding the following constructors for the type formation rule and the term introduction/elimination rules
\begin{alignat*}{3}
  \Pi     &: \implicit{\Gamma}           \; && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
  \lambda &: \implicit{\Gamma, A, B}     \; && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
  \mathsf{app} &: \implicit{\Gamma, A, B}\; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
the computation rules
\[
  \Pi\beta           : \implicit{\Gamma, A, B, t}\; \mathsf{app}\,(\lambda\,t) =^{\Tm\;\Gamma\;A} t
  \qquad\qquad
  \Pi\eta            : \implicit{\Gamma, A, B, t}\; t                         =^{\Tm\;\Gamma\;A} \lambda(\mathsf{app}\,t)
\]
%\begin{alignat*}{5}
%  \Pi\beta           & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t) && =^{\Tm\;\Gamma\;A} && t \\
%  \Pi\eta            & : \implicit{\Gamma, A, B, t}\; && t                      %    && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
%\end{alignat*}
and the structural rules for substitution
\begin{alignat}{5}
  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && \subTy{(\Pi\;A\;B)}{\sigma} && =^{\Ty\;\Gamma} && \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) \label{eq:def-type-subst-8} \\
  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && \subTm{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
\end{alignat}
where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_T}{\color{black}\pi_2\idS}})$.

For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are strict.
we do not need any transport since $[;]_T$ and $[]\Pi$ are strict.
However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $\subTy{\blank}{\blank}$ breaks local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
For example, $\subTy{(\Pi\;A\;B)}{\idS}$ reduces to $\Pi\;A\;(\subTy{B}{\idS \uparrow A})$ where $\idS \uparrow A$ is propositionally but not strictly equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow \subTy{(\Pi\;A\;B)}{\idS} \reduce \Pi\;A\;(\subTy{B}{\idS \uparrow A})$ cannot be joined.

To retain the local confluence, we turn $\sigma \uparrow A$ into a simultaneously defined function:
\begin{alignat}{3}
  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, \subTy{A}{\sigma})\, (\Delta, A) \label{eq:type-of-lifting} \\
\idS                \uparrow A & \reduce \idS \nonumber \\
(\sigma ; \tau)     \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \nonumber \\
\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \nonumber \\
\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow \sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A)\nonumber \\
\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwise} \nonumber
\end{alignat}
where $\sigma^+ \defeq (\pi_1\idS ; \sigma), \pi_2\idS$.
Then, we add another clause for type substitution
\begin{equation*}
  \subTy{(\Pi\,A\,B)}{\sigma} \reduce \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) %\label{eq:def-type-subst-9}
\end{equation*}
and an equality constructor for term substitution without transport:
\begin{alignat}{5}
  \mathsf{[]\lambda} :{} & \sub{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
\end{alignat}

@@

As we are working with a different type theory from \cref{subsec:SC+U}, we  have to again prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$, but the same argument still applies:
\begin{proposition} \label{prop:correctness-3}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
\end{proposition}

Finally, we show that these simultaneously defined functions are coherent.
\begin{restatable}{proposition}{propCoherenceUparrow} %[Coherence]
  For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$ and  any type $A$ and term $t$,   \danger we have
  $\sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, \subTy{A}{\gamma})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A$
      and
      $\subTm{t}{\sigma} =^{\lam{\gamma}\Tm\;\Gamma\;(\subTy{A}{\gamma})}_{\alert{\cc}} \subTm{t}{\tau}$.
    \end{restatable}

@@

We also need to prove coherence of type substitution, i.e., for every equality constructor $\cc : \sigma = \tau$, we have $\subTy{A}{\sigma} = \subTy{A}{\tau}$ .
To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
\[
  \subTy{(\Pi\;A\;B)}{\sigma} = \Pi\;(\subTy{A}{\sigma})\;(\subTy{B}{\sigma\uparrow A}) \qquad
  \subTy{(\Pi\;A\;B)}{\tau} = \Pi\;(\subTy{A}{\tau})\;(\subTy{B}{\tau\uparrow A})
\]
but the induction hypothesis does not give us $\subTy{B}{\sigma\uparrow A} = \subTy{B}{\tau\uparrow A}$.
Instead, we define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ of $\sigma$ by a telescope $\Xi = A_1, \dots, A_n$, given by multiple applications of lifting $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
For each constructor $\cc : \sigma = \tau$, we then show %the identity
$\subTy{A}{\sigma \mathop{\upuparrows} \Xi} = \subTy{A}{\tau \mathop{\upuparrows} \Xi}$
for arbitrary $\Xi$, subsuming the special case $\subTy{A}{\sigma} = \subTy{A}{\tau}$ when $\Xi$ is empty.

The type of telescopes $\Tel : \Ctx \to \Set$ over a `base' context $\Gamma$ are defined inductive-recursively together with a concatenation $\blank \dplus\blank : (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx$ as follows.

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  %\Tel               & :                   && \Ctx \to \Set                                     \\
  \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  %\blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
  \Gamma & \dplus \emptytel && \reduce \Gamma \\
  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
Equipped with telescopes, we can now generalise single lifting with type \eqref{eq:type-of-lifting} to multiple lifting with type
\[
\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
\to \Sub\;(\Gamma \dplus (\subTel{\Xi}{\sigma}))\;(\Delta \dplus \Xi)
\]
which requires us to also simultaneously define telescope substitution $\subTel{\blank}{\blank} :  \Tel\;\Delta \to \Sub\;\Gamma\;\Delta \to \Tel\;\Gamma$.

@@

\begin{minipage}{.5\textwidth}
\begin{alignat*}{3}
  & \subTel{\Xi}{\idS}          && \reduce \Xi \\
  & \subTel{\Xi}{\sigma ; \tau} && \reduce \subTel{(\subTel{\Xi}{\tau})}{\sigma}
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & \idS          && \upuparrows \Xi && \reduce \idS \\
  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
\end{alignat*}
\end{minipage}\vskip\belowdisplayskip\noindent
%whose confluence property is easy to see.
